Binocular integration of chromatic and luminance signals

### 2024-04-07 manuscript.qmd
Added affiliations and abstract

---
title: "Binocular integration of chromatic and luminance signals"
author:
  - Daniel H. Baker
  - Kirralise J. Hansford
  - Federico G. Segala
  - Annie Y. Morsi
  - Rowan J. Huxley
  - Joel T. Martin
  - Maya Rockman
  - Alex R. Wade
format:
  pdf:
    keep-tex: true
  # docx:
  #   toc: false
  #   number-sections: true
# prefer-html: true
pdf-engine: xelatex
# execute:
#   echo: false
bibliography: references.bib
csl: journal-of-vision.csl
header-includes: 
  \usepackage{float} \floatplacement{figure}{H} 
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
---

@@

@@

A further property of 'blob' regions is that they are less strongly orientation-tuned than other regions of V1 [@Horton1981; @Livingstone1984], and biased towards low spatial frequencies [@Edwards1995; @Tootell1988]. Our recent work has investigated binocular combination for flickering discs of luminance, which are DC-balanced across time [@Segala2023]. Steady-state EEG responses from early visual cortex and psychophysical contrast matching data were both consistent with weak interocular suppression when using this stimulus arrangement. This is very different from the 'ocularity invariance' (where binocular and monocular stimuli appear equal) that is well-established when using DC-balanced periodic stimuli such as sine-wave gratings, and implies strong interocular suppression [@Meese2006; @Baker2017; @Moradi2009]. There is additional psychophysical evidence from matching studies that binocular combination can be close to linear for luminance increments [@Levelt1965; @Anstis1998], particularly against a dark background [@Baker2012]. Both structural and functional data therefore imply that binocular combination may differ between spatial and temporal contrast.

The main aim of the present study is to characterise binocular signal combination for chromatic stimuli, and for temporal modulations of luminance. We also aimed to investigate interocular suppression between chromatic and achromatic pathways. We therefore preregistered a series of psychophysical experiments (see: [https://osf.io/3vdga/](https://osf.io/3vdga/)). In Experiment 1 we replicate the four key pedestal masking conditions of @Meese2006 described above for achromatic grating stimuli, and extend this to both red/green and blue/yellow isoluminant chromatic stimuli. In Experiment 2 we explore dichoptic masking within and between these stimuli. Experiment 3 repeats the achromatic condition from the first experiment, but using a temporally modulated disc rather than sine-wave gratings. We take a Bayesian approach to data analysis and modelling; by fitting a hierarchical version of the two-stage gain control model [@Meese2006] we compare posterior parameter distributions to understand how model parameters such as the weight of interocular suppression vary across visual pathways.

>>>>

---
title: "Binocular integration of chromatic and luminance signals"
author:
  - name: Daniel H. Baker$^{1,2,*}$
  - name: Kirralise J. Hansford$^1$
  - name: Federico G. Segala$^1$
  - name: Annie Y. Morsi$^1$
  - name: Rowan J. Huxley$^{1,3}$
  - name: Joel T. Martin$^{1,4}$
  - name: Maya Rockman$^1$
  - name: Alex R. Wade$^{1,2}$
pdf-engine: xelatex
bibliography: references.bib
csl: journal-of-vision.csl
format:
  pdf:
    keep-tex: true
header-includes: 
  \usepackage{float} \floatplacement{figure}{H} 
  \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
---

$^1$Department of Psychology, University of York, UK

$^2$York Biomedical Research Institute, University of York, UK

$^3$School of Psychology, University of Nottingham, UK

$^4$School of Philosophy, Psychology and Language Sciences, University of Edinburgh, UK

$^*$Corresponding author, email: daniel.baker\@york.ac.uk

@@

Much progress has been made in understanding how the brain combines signals from the two eyes. However most of this work has involved achromatic (black and white) stimuli, and it is not clear if the same processes apply in colour-sensitive pathways. In our first experiment, we measured contrast discrimination ('dipper') functions for four key ocular configurations (monocular, binocular, half-binocular and dichoptic), for achromatic, isoluminant red/green and isoluminant blue/yellow sine-wave grating stimuli. We find a similar pattern of results across stimuli, implying equivalently strong interocular suppression within each pathway. Our second experiment measured dichoptic masking within and between pathways using the method of constant stimuli. Masking was strongest within-pathway, and weakest between chromatic and achromatic mechanisms. Finally, we repeated the dipper experiment using temporal luminance modulations, which produced slightly weaker interocular suppression than for spatially modulated stimuli. We interpret our results in the context of a contemporary two-stage model of binocular contrast gain control, implemented here using a hierarchical Bayesian framework. Posterior distributions of the weight of interocular suppression overlapped with a value of 1 for all dipper data sets, and the model captured well the pattern of threshold and slope values from all three experiments.

**Keywords**: *isoluminant*, *dichoptic*, *binocular interactions*, *gain control*, *summation*, *suppression*, *masking*, *psychophysics*

@@

A further property of 'blob' regions is that they are less strongly orientation-tuned than other regions of V1 [@Horton1981; @Livingstone1984], and biased towards low spatial frequencies [@Edwards1995; @Tootell1988]. Our recent work [@Segala2023] has investigated binocular combination for flickering discs of luminance, which are DC-balanced across time (i.e. the time-averaged luminance is equal to the background). Steady-state EEG responses from early visual cortex and psychophysical contrast matching data were both consistent with weak interocular suppression when using this stimulus arrangement. This is very different from the 'ocularity invariance' (where binocular and monocular stimuli appear equal) that is well-established when using DC-balanced periodic stimuli such as sine-wave gratings, and implies strong interocular suppression [@Meese2006; @Baker2017; @Moradi2009]. There is additional psychophysical evidence from matching studies that binocular combination can be close to linear for luminance increments [@Levelt1965; @Anstis1998], particularly against a dark background [@Baker2012]. Both structural and functional data therefore imply that binocular combination may differ between spatial and temporal contrast.

The main aim of the present study was to characterise binocular signal combination for chromatic stimuli, and for temporal modulations of luminance. We also aimed to investigate interocular suppression between chromatic and achromatic pathways. We therefore preregistered a series of psychophysical experiments (see: [https://osf.io/3vdga/](https://osf.io/3vdga/)). In Experiment 1 we replicate the four key pedestal masking conditions of @Meese2006 described above for achromatic grating stimuli, and extend this to both red/green and blue/yellow isoluminant chromatic stimuli. In Experiment 2 we explore dichoptic masking within and between these stimuli. Experiment 3 repeats the achromatic condition from the first experiment, but using a temporally modulated disc rather than sine-wave gratings. We take a Bayesian approach to data analysis and modelling; by fitting a hierarchical version of the two-stage gain control model [@Meese2006] we compare posterior parameter distributions to understand how model parameters such as the weight of interocular suppression vary across visual pathways.

### 2024-04-07 manuscript.tex
Added affiliations and abstract

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\usepackage{float} \floatplacement{figure}{H} \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Binocular integration of chromatic and luminance signals},
  pdfauthor={Daniel H. Baker; Kirralise J. Hansford; Federico G. Segala; Annie Y. Morsi; Rowan J. Huxley; Joel T. Martin; Maya Rockman; Alex R. Wade},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Binocular integration of chromatic and luminance signals}
\author{Daniel H. Baker \and Kirralise J. Hansford \and Federico G.
Segala \and Annie Y. Morsi \and Rowan J. Huxley \and Joel T.
Martin \and Maya Rockman \and Alex R. Wade}
\date{}

\begin{document}
\maketitle
\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

@@

A further property of `blob' regions is that they are less strongly
orientation-tuned than other regions of V1 (Horton \& Hubel, 1981;
Livingstone \& Hubel, 1984), and biased towards low spatial frequencies
(Edwards et al., 1995; Tootell et al., 1988). Our recent work has
investigated binocular combination for flickering discs of luminance,
which are DC-balanced across time (Segala et al., 2023). Steady-state
EEG responses from early visual cortex and psychophysical contrast
matching data were both consistent with weak interocular suppression
when using this stimulus arrangement. This is very different from the
`ocularity invariance' (where binocular and monocular stimuli appear
equal) that is well-established when using DC-balanced periodic stimuli
such as sine-wave gratings, and implies strong interocular suppression
(Baker \& Wade, 2017; Meese et al., 2006; Moradi \& Heeger, 2009). There
is additional psychophysical evidence from matching studies that
binocular combination can be close to linear for luminance increments
(Anstis \& Ho, 1998; Levelt, 1965), particularly against a dark
background (Baker et al., 2012). Both structural and functional data
therefore imply that binocular combination may differ between spatial
and temporal contrast.

The main aim of the present study is to characterise binocular signal
combination for chromatic stimuli, and for temporal modulations of
luminance. We also aimed to investigate interocular suppression between
chromatic and achromatic pathways. We therefore preregistered a series
of psychophysical experiments (see: \url{https://osf.io/3vdga/}). In
Experiment 1 we replicate the four key pedestal masking conditions of
Meese et al. (2006) described above for achromatic grating stimuli, and
extend this to both red/green and blue/yellow isoluminant chromatic
stimuli. In Experiment 2 we explore dichoptic masking within and between
these stimuli. Experiment 3 repeats the achromatic condition from the
first experiment, but using a temporally modulated disc rather than
sine-wave gratings. We take a Bayesian approach to data analysis and
modelling; by fitting a hierarchical version of the two-stage gain
control model (Meese et al., 2006) we compare posterior parameter
distributions to understand how model parameters such as the weight of
interocular suppression vary across visual pathways.

>>>>

\usepackage{booktabs}
\usepackage{longtable}
\usepackage{array}
\usepackage{multirow}
\usepackage{wrapfig}
\usepackage{float}
\usepackage{colortbl}
\usepackage{pdflscape}
\usepackage{tabu}
\usepackage{threeparttable}
\usepackage{threeparttablex}
\usepackage[normalem]{ulem}
\usepackage{makecell}
\usepackage{xcolor}
\KOMAoption{captions}{tableheading}
\usepackage{float} \floatplacement{figure}{H} \newcommand{\beginsupplement}{\setcounter{table}{0}  \renewcommand{\thetable}{A\arabic{table}} \setcounter{figure}{0} \renewcommand{\thefigure}{A\arabic{figure}}}
\makeatletter
\makeatother
\makeatletter
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\AtBeginDocument{%
\ifdefined\contentsname
  \renewcommand*\contentsname{Table of contents}
\else
  \newcommand\contentsname{Table of contents}
\fi
\ifdefined\listfigurename
  \renewcommand*\listfigurename{List of Figures}
\else
  \newcommand\listfigurename{List of Figures}
\fi
\ifdefined\listtablename
  \renewcommand*\listtablename{List of Tables}
\else
  \newcommand\listtablename{List of Tables}
\fi
\ifdefined\figurename
  \renewcommand*\figurename{Figure}
\else
  \newcommand\figurename{Figure}
\fi
\ifdefined\tablename
  \renewcommand*\tablename{Table}
\else
  \newcommand\tablename{Table}
\fi
}
\@ifpackageloaded{float}{}{\usepackage{float}}
\floatstyle{ruled}
\@ifundefined{c@chapter}{\newfloat{codelisting}{h}{lop}}{\newfloat{codelisting}{h}{lop}[chapter]}
\floatname{codelisting}{Listing}
\newcommand*\listoflistings{\listof{codelisting}{List of Listings}}
\makeatother
\makeatletter
\@ifpackageloaded{caption}{}{\usepackage{caption}}
\@ifpackageloaded{subcaption}{}{\usepackage{subcaption}}
\makeatother
\makeatletter
\makeatother
\ifLuaTeX
  \usepackage{selnolig}  % disable illegal ligatures
\fi
\IfFileExists{bookmark.sty}{\usepackage{bookmark}}{\usepackage{hyperref}}
\IfFileExists{xurl.sty}{\usepackage{xurl}}{} % add URL line breaks if available
\urlstyle{same} % disable monospaced font for URLs
\hypersetup{
  pdftitle={Binocular integration of chromatic and luminance signals},
  pdfauthor={Daniel H. Baker\^{}\{1,2,*\}; Kirralise J. Hansford\^{}1; Federico G. Segala\^{}1; Annie Y. Morsi\^{}1; Rowan J. Huxley\^{}\{1,3\}; Joel T. Martin\^{}\{1,4\}; Maya Rockman\^{}1; Alex R. Wade\^{}\{1,2\}},
  colorlinks=true,
  linkcolor={blue},
  filecolor={Maroon},
  citecolor={Blue},
  urlcolor={Blue},
  pdfcreator={LaTeX via pandoc}}

\title{Binocular integration of chromatic and luminance signals}
\author{Daniel H. Baker\(^{1,2,*}\) \and Kirralise J.
Hansford\(^1\) \and Federico G. Segala\(^1\) \and Annie Y.
Morsi\(^1\) \and Rowan J. Huxley\(^{1,3}\) \and Joel T.
Martin\(^{1,4}\) \and Maya Rockman\(^1\) \and Alex R. Wade\(^{1,2}\)}
\date{}

\begin{document}
\maketitle
\(^1\)Department of Psychology, University of York, UK

\(^2\)York Biomedical Research Institute, University of York, UK

\(^3\)School of Psychology, University of Nottingham, UK

\(^4\)School of Philosophy, Psychology and Language Sciences, University
of Edinburgh, UK

\(^*\)Corresponding author, email: daniel.baker@york.ac.uk

\hypertarget{abstract}{%
\section{Abstract}\label{abstract}}

Much progress has been made in understanding how the brain combines
signals from the two eyes. However most of this work has involved
achromatic (black and white) stimuli, and it is not clear if the same
processes apply in colour-sensitive pathways. In our first experiment,
we measured contrast discrimination (`dipper') functions for four key
ocular configurations (monocular, binocular, half-binocular and
dichoptic), for achromatic, isoluminant red/green and isoluminant
blue/yellow sine-wave grating stimuli. We find a similar pattern of
results across stimuli, implying equivalently strong interocular
suppression within each pathway. Our second experiment measured
dichoptic masking within and between pathways using the method of
constant stimuli. Masking was strongest within-pathway, and weakest
between chromatic and achromatic mechanisms. Finally, we repeated the
dipper experiment using temporal luminance modulations, which produced
slightly weaker interocular suppression than for spatially modulated
stimuli. We interpret our results in the context of a contemporary
two-stage model of binocular contrast gain control, implemented here
using a hierarchical Bayesian framework. Posterior distributions of the
weight of interocular suppression overlapped with a value of 1 for all
dipper data sets, and the model captured well the pattern of threshold
and slope values from all three experiments.

\textbf{Keywords}: \emph{isoluminant}, \emph{dichoptic}, \emph{binocular
interactions}, \emph{gain control}, \emph{summation},
\emph{suppression}, \emph{masking}, \emph{psychophysics}

@@

A further property of `blob' regions is that they are less strongly
orientation-tuned than other regions of V1 (Horton \& Hubel, 1981;
Livingstone \& Hubel, 1984), and biased towards low spatial frequencies
(Edwards et al., 1995; Tootell et al., 1988). Our recent work (Segala et
al., 2023) has investigated binocular combination for flickering discs
of luminance, which are DC-balanced across time (i.e.~the time-averaged
luminance is equal to the background). Steady-state EEG responses from
early visual cortex and psychophysical contrast matching data were both
consistent with weak interocular suppression when using this stimulus
arrangement. This is very different from the `ocularity invariance'
(where binocular and monocular stimuli appear equal) that is
well-established when using DC-balanced periodic stimuli such as
sine-wave gratings, and implies strong interocular suppression (Baker \&
Wade, 2017; Meese et al., 2006; Moradi \& Heeger, 2009). There is
additional psychophysical evidence from matching studies that binocular
combination can be close to linear for luminance increments (Anstis \&
Ho, 1998; Levelt, 1965), particularly against a dark background (Baker
et al., 2012). Both structural and functional data therefore imply that
binocular combination may differ between spatial and temporal contrast.

The main aim of the present study was to characterise binocular signal
combination for chromatic stimuli, and for temporal modulations of
luminance. We also aimed to investigate interocular suppression between
chromatic and achromatic pathways. We therefore preregistered a series
of psychophysical experiments (see: \url{https://osf.io/3vdga/}). In
Experiment 1 we replicate the four key pedestal masking conditions of
Meese et al. (2006) described above for achromatic grating stimuli, and
extend this to both red/green and blue/yellow isoluminant chromatic
stimuli. In Experiment 2 we explore dichoptic masking within and between
these stimuli. Experiment 3 repeats the achromatic condition from the
first experiment, but using a temporally modulated disc rather than
sine-wave gratings. We take a Bayesian approach to data analysis and
modelling; by fitting a hierarchical version of the two-stage gain
control model (Meese et al., 2006) we compare posterior parameter
distributions to understand how model parameters such as the weight of
interocular suppression vary across visual pathways.
